Streaking and Clutching in Round-Based First Person Shooter Esports

### 2022-07-17 presentation.qmd
analytic formulas for series outcomes and streaks

::: fragment
-   Plant/Defuse in Valorant
    -   Data from Valorant professional league, updated through January 1, 2022
    -   5v5
    -   First to 13 round wins
    -   One team plays offense in 12 consecutive rounds before swapping sides
    -   Must win by 2 rounds
:::

@@

```{r}
#| label: p_series_outcome_prop_over_expected
prior_clinching_round_win_prop <- rounds |> 
  filter(round == (n_rounds - 1L)) |> 
  count(n_rounds, win_series, won_prior_round = win_round) |> 
  mutate(
    across(win_series, ~ifelse(.x, 'win', 'loss'))
  ) |> 
  pivot_wider(
    names_from = win_series,
    values_from = n,
    values_fill = 0L
  ) |> 
  filter(won_prior_round) |> 
  select(-won_prior_round) |> 
  mutate(
    win_series_prop = win / (win + loss),
    e_win_series_prop = (6L - 1L) / (n_rounds - 1L)
  )

n_rounds_max <- 11
n_sims <- 100000
set.seed(42)
w <- sample(c(0, 1), size = n_rounds_max * n_sims, replace = TRUE)

sim_rounds <- tibble(
  i = rep(1:n_sims, each = n_rounds_max),
  r = rep(1:n_sims, times = n_rounds_max),
  w = w
) |> 
  group_by(i) |> 
  mutate(
    cumu_w = cumsum(w),
    cumu_l = cumsum(w == 0)
  ) |> 
  ungroup() |> 
  mutate(
    cumu_wl_max = ifelse(cumu_w > cumu_l, cumu_w, cumu_l)
  ) |> 
  filter(cumu_wl_max <= 6)

sim_rounds_to_drop <- anti_join(
  sim_rounds |> 
    filter(cumu_wl_max == 6L),
  sim_rounds |> 
    filter(cumu_wl_max == 6L) |> 
    group_by(i) |> 
    slice_min(r, n = 1) |> 
    ungroup()
)

sim_rounds <- sim_rounds |> 
  anti_join(
    sim_rounds_to_drop
  )

## always from offensive perspective
e_series_streaks <- sim_rounds |> 
  inner_join(
    sim_rounds |> 
      filter(cumu_wl_max == 6) |> 
      mutate(
        zeros_win = cumu_l == cumu_wl_max
      ) |> 
      select(i, zeros_win)
  ) |> 
  mutate(
    across(w, ~ifelse(zeros_win, abs(1 - .x), .x))
  ) |> 
  group_by(i) |> 
  summarize(
    cumu_w = max(cumu_w),
    cumu_l = max(cumu_l),
    ws = paste0(w, collapse = '')
  ) |> 
  ungroup() |> 
  mutate(
    wins = ifelse(cumu_w == 6, cumu_w, cumu_l),
    losses = ifelse(cumu_w == 6, cumu_l, cumu_w)
  ) |> 
  unite(
    record, wins, losses, sep = '-'
  ) |> 
  count(record, ws, sort = TRUE) |> 
  mutate(prop = n / sum(n))

actual_series_streaks <- rounds |> 
  filter(win_series) |> 
  mutate(across(win_round, as.integer)) |> 
  group_by(year, event, series) |> 
  summarize(
    wins = max(cumu_w),
    losses = max(cumu_l),
    ws = paste0(win_round, collapse = '')
  ) |> 
  ungroup() |> 
  unite(
    record, wins, losses, sep = '-'
  ) |> 
  count(record, ws, sort = TRUE) |> 
  mutate(prop = n / sum(n))

series_streaks <- full_join(
  actual_series_streaks |> 
    rename_with(~sprintf('%s_actual', .x), c(n, prop)),
  e_series_streaks |> 
    rename_with(~sprintf('%s_expected', .x), c(n, prop))
) |> 
  mutate(
    prop_diff = prop_actual - prop_expected
  ) |> 
  arrange(desc(abs(prop_diff)))
# series_streaks |> filter(is.na(n_actual))

## 6-0, 6-1, and some combos of 6-5 occur more often than expected
series_streak_props <- series_streaks |> 
  filter(!is.na(n_actual)) |> 
  mutate(total_actual = sum(n_actual)) |> 
  mutate(
    p = vectorized_prop_test(n_actual, total_actual, n_expected, !!n_sims)
  ) |> 
  select(-total_actual) |> 
  arrange(p$p.value)

## 6-0 and 6-1 occur more often than expected
series_outcome_prop <- full_join(
  actual_series_streaks |> 
    group_by(record) |> 
    summarize(
      across(n, sum)
    ) |> 
    mutate(prop = n / sum(n)) |> 
    rename_with(~sprintf('%s_actual', .x), c(n, prop)),
  e_series_streaks |> 
    group_by(record) |> 
    summarize(
      across(n, sum)
    ) |> 
    mutate(prop = n / sum(n)) |> 
    rename_with(~sprintf('%s_expected', .x), c(n, prop))
) |> 
  mutate(
    prop_diff = prop_actual - prop_expected,
    total_actual = sum(n_actual),
    n_expected_downsampled = round(prop_expected * total_actual),
    p = vectorized_prop_test(n_actual, total_actual, n_expected_downsampled, total_actual)
  ) |> 
  select(-total_actual) |> 
  arrange(p$p.value)

long_series_outcome_prop <- series_outcome_prop |>
  unnest(p) |>
  select(
    record,
    n = n_actual,
    `3` = prop_actual,
    `2` = prop_expected,
    `1` = prop_diff
  ) |> 
  pivot_longer(
    matches('[123]'),
    names_to = 'outcome_idx',
    values_to = 'value'
  ) |> 
  mutate(across(outcome_idx, as.integer))

@@

::: fragment
-   Round win % after a streak of 3 round wins:
    -   Expected: **`r 100 * round(prob_n6_k3, 3)`%**
    -   Actual: **`r 100 * round(round_win_prop_after_b2b2b_wins$prop, 3)`%** (`r round_win_prop_after_b2b2b_wins$n` / `r round_win_prop_after_b2b2b_wins$total`)
:::

@@

## What is the optimal strategy?

>>>>

::: fragment
-   Plant/Defuse in Valorant
    -   Valorant professional league, updated through January 1, 2022
    -   5v5
    -   First to 13 round wins
    -   One team plays offense in 12 consecutive rounds before swapping sides
    -   Must win by 2 rounds
:::

@@

```{r}
#| label: p_series_outcome_prop_over_expected
# https://content.iospress.com/articles/journal-of-sports-analytics/jsa200422
prob_of_series_lasting_n_games <- function(n, p = 0.5, max_rounds = 11) {
  s <- ceiling(max_rounds / 2)
  (factorial(n - 1) / (factorial(s - 1) * factorial(n - s))) * (p^s * (1 - p)^(n - s) + p^(n - s) * (1 - p)^s)
}

possibilities_of_outcomes <- function(m, n) {
  factorial(m + n) / (factorial(m) * factorial(n))
}

actual_series_streaks <- rounds |> 
  filter(win_series) |> 
  mutate(across(win_round, as.integer)) |> 
  group_by(year, event, series) |> 
  summarize(
    wins = max(cumu_w),
    losses = max(cumu_l),
    ws = paste0(win_round, collapse = '')
  ) |> 
  ungroup() |> 
  mutate(n_rounds = wins + losses) |> 
  unite(
    record, wins, losses, sep = '-'
  ) |> 
  count(n_rounds, record, ws, sort = TRUE) |> 
  mutate(prop = n / sum(n))

possibilities <- tibble(
  n_rounds = 6L:11L
) |> 
  mutate(
    series_prop = map_dbl(n_rounds, prob_of_series_lasting_n_games),
    possibilities = map2_dbl(6, n_rounds - 6, possibilities_of_outcomes)
  ) |> 
  transmute(
    n_rounds,
    series_prop,
    record = sprintf('6-%d', n_rounds - 6),
    prop = series_prop / possibilities
  )

series_streaks <- inner_join(
  actual_series_streaks |> 
    rename_with(~sprintf('%s_actual', .x), c(n, prop)),
  possibilities |> 
    rename_with(~sprintf('%s_expected', .x), prop),
  by = 'n_rounds'
) |> 
  mutate(
    prop_diff = prop_actual - prop_expected
  ) |> 
  arrange(desc(abs(prop_diff)))
# series_streaks |> filter(is.na(n_actual))

## 6-0, 6-1, and some combos of 6-5 occur more often than expected
series_streak_props <- series_streaks |> 
  filter(!is.na(n_actual)) |> 
  mutate(
    prop_diff = prop_actual - prop_expected,
    total_actual = sum(n_actual),
    n_expected = round(prop_expected * total_actual),
    p = vectorized_prop_test(n_actual, total_actual, n_expected, total_actual)
  ) |> 
  select(-total_actual) |> 
  unnest_wider(p) |> 
  arrange(p.value)

## 6-0 and 6-1 occur more often than expected
series_outcome_prop <- full_join(
  actual_series_streaks |> 
    group_by(record) |> 
    summarize(
      across(n, sum)
    ) |> 
    mutate(prop = n / sum(n)) |> 
    rename_with(~sprintf('%s_actual', .x), c(n, prop)),
  possibilities |> 
    select(record, prop_expected = series_prop),
  by = 'record'
) |> 
  mutate(
    prop_diff = prop_actual - prop_expected,
    total_actual = sum(n_actual),
    n_expected = round(prop_expected * total_actual),
    p = vectorized_prop_test(n_actual, total_actual, n_expected, total_actual)
  ) |> 
  select(-total_actual) |> 
  unnest_wider(p) |> 
  arrange(p.value)

long_series_outcome_prop <- series_outcome_prop |>
  select(
    record,
    n = n_actual,
    `3` = prop_actual,
    `2` = prop_expected,
    `1` = prop_diff
  ) |> 
  pivot_longer(
    matches('[123]'),
    names_to = 'outcome_idx',
    values_to = 'value'
  ) |> 
  mutate(across(outcome_idx, as.integer))

@@

-   Round win % after a streak of 3 round wins:
    -   Expected: **`r 100 * round(prob_n6_k3, 3)`%**
    -   Actual: **`r 100 * round(round_win_prop_after_b2b2b_wins$prop, 3)`%** (`r round_win_prop_after_b2b2b_wins$n` / `r round_win_prop_after_b2b2b_wins$total`)

@@

## Is planting at one site better than the other?
